Management induced changes of soil organic carbon on global croplands

### 2020-11-05 SOCbudgetPaper/discussion.Rmd
after spell check

- The natural land representation and therefor also SOC initialization is lacking proper parametrization of both decay behaviour and natural input (e.g. nitrogen and lignin content of litterfall). As all results are valued against these potential natural SOC the absolute values and also the changes over time have to be used with caution. Especially in boreal zone and in very dry regions the natural land representation might be far off, due to the focus for the parametrization for cropped areas. 
The good agreement of the stock change factors $F^{SCF}$ with the Tier 1 default values might serve as a hint that the natural land representation is not totally off. Boreal and very dry zone might be uncropped, leaving the results on SOC changes on cropland untouch from the bad borreal land representation.

- This study as the IPCC guidelines suggested has limited her focus to the first 30 cm of the soil profile, leaving changes in the subsoil unnoticed. Nevertheless studies (see Don on tillage) have shown, the subsoil to be a game changer in evaluating total SOC losses or gains for no-tillage systems. It has been argued that for intensivly tilled soils, subsoil SOC is increasing due to the import of carbon rich topsoil to deeper soil layers. Following these argumentation SOC stocks in croplands might even be underestimated.

- As pointed out by Keel (2017), Smith (2019) the results might be highly sensitive to carbon input calculations more precisly to below and above ground residue carbon estimates derived from harvested quantities. It has been questioned, that below ground residues might increase with a fixed root:shot ratio (maybe specifically in high end farming systems (?)). Following this argumentation SOC results shown in this study might especially in high-yielding farming system (europe etc.) overestimate actual SOC stocks. (In general the sensitivity to and the uncertainty of certain parameters might be very high.)

<!--Tasks: 
Find better natural land representation and initalization. Using LPJmL values might be complex since the split up into SOC subpools is needed. Getting better parametrization for the CENTURY based approach for decay behaviour and better values for nitrogen and lignin of natural litter fall.  -->

\newpage

Smaller points and shortcommings:

- Fertilizer interaction is not included here by accounting for additional N supply that would alter C:N ratio of the carbon inputs. Tier 2 steady-state method is neglecting fertilizer application, however we would have fertilizer ammounts at hand to include them, if proper representation of fertilizer within the method would be possible to add.

- Pasture dynamics are neglected and treated as natural vegetation, which might be -- looking on pasture degradation due to overgrazing -- oversimplified for some spots, but is inline with assumption on pasture SOC stocks done before (see Tier 1 IPCC). (Note that also manure excreted to pastures is neglected within these analysis, since we focus purely on cropland dynamics.)

- No tillage adaptaion is neglected on cropland due to less common adoption of no tillage and conservation agriculture. Pastures are assumed to be not tilled at all (propably only heavy managed pastures are tilled with some rotation)

- Irrigated areas are not crop specific and irrigation is not restricted to growing periods (since it is very complex to calculate average growing periods). Crop specific growing periods might be possible using LPJmL data.

- flooded rice area are not represented correctly as parameterization does not hold true for flooded conditions.

@@

- The disaggregation of manure to uild-up areas (in the case of extensive monograstrics) is leeding to a lot of displaced manure (?) that is cut off

- It is known that there are mismatches between FAO statistics and LUH areas. As far as possibles there were harmonized within this study.

- "The Tier 2 method does not simulate C change but simply calculates an annual C stock change from the
current C stock to the future steady-state soil C stock calculated based on current conditions." Leading to the fact that our total stock results are highly uncertained.

>>>>

## Including agricultural management data changes the sign of the trend

This study provides an analysis of historic SOC stock changes on cropland. We determine the SOC trends on cropland compared to a counterfactual scenario with a world under natural vegetation under identical historical climatic conditions ($SOC_{natveg}$). Our results show that human activity lead to cumulative SOC emission of around 37 GtC in 2010. Whereas recent modelling estimates of global SOC emissions indicate an ongoing increase of SOC emissions (Sanderman et al, Pugh et al.), our study indicates that the global SOC gap is slowly closing due to improved management.

According to Sanderman et al. cumulative SOC emissions since the beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil, with half of this value attributed to grazing. Our estimate of 37 GtC in 2010 for cropland emissions only is twice as high as Sanderman et al. estimations. However, there are large uncertainties in modelling SOC at the global scale, and the Sanderman et al. results have been pointed out as potentially conservatively low compared with experimental results, leaving our results in an aceptable range.

Furthermore, Sanderman et al. modelled historical trends based on agricultural land expansion without considering SOC variations due to different management systems at all. Pugh et al. considered management effects like tillage and residue recycling in a static way, but neither changes over time nor alignment to observed historical data like yield-levels or no-tillage areas were taken into account. The study moreover concludes that crop productivity gains (increasing yield levels by 18%) do not lead to a substantial decrease in SOC emissions, accounting for only less than 1% change in SOC emissions.

Our study for the first time uses a dynamic management dataset as driver for SOC dynamics. We show that the moderate global cropland expansion of around 11% between 1974 and 2010 and the resulting depletion of SOC stocks in converted cropland has been outweighed by improved agricultural yields and practices. This challenges Pugh et al.'s findings of only small effects due to improved practices.
Moreover, our sensitivity analysis indicates that (1) recycled residue biomass increases driven by higher yields as well as higher recycling rates play the most important role, followed by (2) improved manure recycling (e.g. due to improved animal waste management systems) and (3) the adoption of no tillage practices.

Modelling management effects on the global scale comes however with parametric and structural uncertainties. As pointed out by Keel (2017) and Smith (2019), carbon input calculations are highly sensitive to the choice of allometric functions
determining below and above ground residue estimates from harvested quantities. Keel et al. question whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shifts allometries which might not be reflected in outdated data sources. Following this line of argumentation, SOC results calculated in this study may overestimate actual SOC stocks due to an overestimation in residue biomass, for which the ratio compared to the harvested organ is normally reduced due to breeding. However, with an evaluation of management effects (see below), there is no indication for a systematic overrating of residue biomass. Rather, it is more likely that we are still missing carbon inputs to the soil.

## Modeled management effect in line with default IPCC assumptions

To validate our modelled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors (cite) which are based on measurement data for croplands (see \@ref(tab:SCFtable)). To allow for comparison, we aggregate our stock change factors to the four IPCC climate zones.

```{r captionSCFtable, echo=FALSE}
captionSCFtable <- "This table shows different estimates of stock change factors $F_{SCF}$, defined as the SOC stock of managed croplands relative to the counterfactual SOC stock of undisturbed natural vegetation. We compare average values for four IPCC climate zone classifications. IPCC 2006 and 2019 default factors (medium input) as well as altered factors for low input of organic matter without any other subsystem consideration are compared to results for 1990 and 2010 of this study (SOC budget).}"
```

```{r SCFtable, out.width = "16cm", fig.cap = captionSCFtable, echo=FALSE}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/TableSCF_comparison.png")
```

Our estimates correspond very well to the default stock-change factors used for the tier 1 estimation of IPCC, 2006. For the tropical regions the assumptions changed notably from the guidelines in 2006 to the update in 2019, leaving our results too low in comparison with IPCC, 2019. Considering yield gaps in mainly developing regions in the tropics the default assumption of medium input systems might be an overestimation of actual SOC state. The additional effect of considering low inputs of biomass in tropical regions can however not explain the full mismatch to IPCC 2019 values but account for 14-21% of it.

With regard to the time trend, our study shows the substantial impact of changing management factors on the development of $F_{SCF}$.

## SOC stocks inline with literature

The worlds SOC stock and its changes are highly uncertain (cite), which is seen in the wide range of global SOC stock estimates (see \@ref(tab:SOCtable)).

```{r captionSOCtable, echo=FALSE}
captionSOCtable <- "Modelled as well as data based estimation for global SOC stock in GtC for the first 30 cm of soil aggregated over all land area. Note that SoilGrids, GSOC and WISE do not consider changes over time and rely on soil profile data gather over a long period of time, which makes it hard to pinpoint a specific year for these SOC estimations. In this context they will be compared to modelled data (LPJmL4, this study) for the year 2010."
```

```{r SOCtable, out.width = "8cm", fig.cap = captionSOCtable, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/TableSOC_comparison.png")
```

The global estimates of SOC stock from this study are on the lower end compared to other modelled results or more data driven estimates. Looking at regional results (fig. SX in supplement), our estimates turn out to be in good agreement for most regions, with the largest deviations for boreal areas. Considering that the model was parametrized for croplands, these mismatches are not suprising as the temperature effects on decomposition are fundamentally different for permafrost soils. To avoid that this bias influences our results, our study focuses exclusively on cropland soils, excluding most of the boreal zone. Moreover, when focussing on SOC changes on cropland, pristine natural vegetated areas without human land management under the same historic climatic conditions cancel out in the calculation of SOC emissions.

Our estimates for total SOC stocks of the world, as well as our SOC initialization, are dominated by the representation of natural lands and pastures, which are however only estimated in a basic manner. For example, we do not have a differentiated parametrization of nitrogen and lignin content of litterfall for woody and grass type biomes. This leaves carbon inputs and decay behavior for natural land and pastures rather uncertain. The absolute values of stocks and emissions from land-use change therefore have to be used with caution. Especially in less forested areas the natural land representation higher uncertainty may arise from the parametrization assumptions of natural litterfall.

Nevertheless, total SOC stocks are in a reasonable range and results for SOC stock changes in relative or absolute numbers are not altered by pristine vegetated areas at all. The SOC gap from changed management is only affected by the natural vegetation representation in the case of land use transitions. In this case, former natural land SOC stocks decline. However, as pointed out before, due to low cropland expansion rates, land use change-related SOC declines are outweighed by the general trend of increasing SOC stocks on existing cropland due to management improvements.
We conducted a sensitivity analysis for a wide range of possible parameter combinations for lignin and nitrogen parametrization of natural litterfall, which tend to change the global SOC stock substantially for different choices
(see appendix). This shows that the general trend of decreasing SOC gap is not altered even under very high estimates for natural SOC stocks.

## Important shortcomings --> move to appendix

Smaller points and shortcomings:

-This study as the IPCC guidelines suggested has limited its focus to the first 30 cm of the soil profile, leaving changes in the subsoil unnoticed. Nevertheless, studies (see Don on tillage) have shown the subsoil to be very influential in evaluating total SOC losses or gains for no-tillage systems. It has been argued that for intensively tilled soils, subsoil SOC increases due to the import of carbon rich topsoil to deeper soil layers. Following this argumentation, SOC stocks in croplands might even be underestimated.

- Fertilizer interaction is not included here by accounting for additional N supply that would alter the C:N ratio of the carbon inputs. For the IPCC Tier 2 steady-state method, fertilizer application is neglected; However we would have fertilizer amounts available to include, if proper representation of fertilizer within the method would be possible to add.

- Pasture dynamics are neglected and treated as natural vegetation, which may be -- especially with regard to pasture degradation due to overgrazing -- oversimplified for some areas, but is in line with assumptions on previous pasture SOC stocks (see Tier 1 IPCC). (Note that also manure excreted to pastures is neglected within this analysis, since we focus purely on cropland dynamics.)

- No-tillage practices are neglected on cropland due to the less common adoption of no tillage and conservation agriculture. Pastures are assumed to be not tilled at all (probably only heavily managed pastures are tilled with some rotation)

- Irrigated areas are not crop specific and irrigation is not restricted to growing periods (since it is very complex to calculate average growing periods). Crop specific growing periods may be possible using LPJmL data.

- flooded rice area are not represented correctly as the parametrization does not hold true for flooded conditions.

@@

- The disaggregation of manure to build-up areas (in the case of extensive monogastric) is leading to a lot of displaced manure (?) that is cut off

- It is known that there are mismatches between FAO statistics and LUH areas. As far as possible these were harmonized within this study.

- "The Tier 2 method does not simulate C change but simply calculates an annual C stock change from the
current C stock to the future steady-state soil C stock calculated based on current conditions." This leads to the fact that our total stock results are highly uncertain.
